---
title: "sobol_assignment"
format: html
editor: visual
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Key takeaways from the Uranium In Ground Water Model paper

The sensitivity analysis shows uranium concentration and water intake are key factors affecting health risks from groundwater. Managers can now focus monitoring on these specific variables instead of tracking everything. With children facing higher drinking risks and adults more affected by skin contact, targeted protection strategies can be developed for different age groups. This focused approach makes better use of limited resources in contaminated areas and improves public health outcomes overall.

# Part 2: Repeating the sensitivity analysis 

```{r}
library(sensitivity)
library(tidyverse)
library(gridExtra)
library(purrr)
library(ggpubr)
library(here)

# Load conductance model
source(here("R", "Catm.R"))
```

```{r}
# Define number of samples
np <- 1000

# Generate input values using specified distributions for each parameter
k_o <- rnorm(np, mean = 0.1, sd = 0.1 * 0.01)
k_d <- rnorm(np, mean = 0.7, sd = 0.7 * 0.01)
v <- rnorm(np, mean = 300, sd = 50)
height <- runif(np, min = 3.5, max = 5.5)

# Combine into a single data frame for X1
X1 <- cbind.data.frame(k_o, k_d, v, height)

# Repeat process to create independent sample set X2
k_o <- rnorm(np, mean = 0.1, sd = 0.1 * 0.01)
k_d <- rnorm(np, mean = 0.7, sd = 0.7 * 0.01)
v <- rnorm(np, mean = 300, sd = 50)
height <- runif(np, min = 3.5, max = 5.5)

X2 <- cbind.data.frame(k_o, k_d, v, height)

# Set up Sobol sensitivity object
sens_Catm_Sobol <- sobolSalt(model = NULL, X1, X2, nboot = 100)

```

```{r}
# Extract full Sobol parameter set and run the model
parms <- as.data.frame(sens_Catm_Sobol$X)
colnames(parms) <- colnames(X1)

# Use pmap to apply model across all parameter combinations
res <- pmap_dbl(parms, Catm)

# Return results to Sobol object for analysis
sens_Catm_Sobol <- sensitivity::tell(sens_Catm_Sobol, res, res.names = "ga")

```

```{r}
# Combine parameters and outputs into one data frame
both <- cbind.data.frame(parms, gs = sens_Catm_Sobol$y)

# Create histogram showing variation in conductance values
plot1 <- ggplot(both, aes(x = gs)) +
  geom_histogram() +
  geom_vline(xintercept = mean(both$gs), col = "lightpink")

print(plot1)

```
```{r}
# Plot conductance vs. windspeed, colored by vegetation height
plot2.1 <- ggplot(both, aes(v, gs, col = height)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "Windspeed")

# Plot conductance vs. height, colored by windspeed
plot2.2 <- ggplot(both, aes(height, gs, col = v)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "Vegetation Height")

print(plot2.1)
print(plot2.2)

```


```{r}
# Assign readable row names for indices
row.names(sens_Catm_Sobol$S) <- colnames(parms)
row.names(sens_Catm_Sobol$T) <- colnames(parms)

# Display first-order (main effect) Sobol indices
sens_Catm_Sobol$S

# Display total-order Sobol indices
sens_Catm_Sobol$T

# Print complete summary of Sobol object
print(sens_Catm_Sobol)

```

Our results show that both wind and how tall the plants are affect conductance. This is different from what we saw in class, where these didn't matter as much. It looks like shorter plants might let more wind through, which increases the effect. Our model responds to both, but wind seems to have the biggest impact
